---
title: "Introduction to clustVarACC: Variable Clustering with R6 Classes"
author: "Cyrille PECNIK, Constantin REY-COQUAIS, Anne-Camille VIAL"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to clustVarACC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```

## Overview

The **clustVarACC** package provides three R6-based algorithms for clustering variables (rather than observations). Variable clustering is useful for:

- **Reducing multicollinearity** in regression models
- **Feature engineering**: creating representative variables for each group
- **Exploratory data analysis**: understanding variable relationships and structure
- **Dimension reduction**: grouping redundant variables together

The package includes:

1. **ClustVarKMeans**: K-means clustering for numeric variables
2. **ClustVarHAC**: Hierarchical Agglomerative Clustering for numeric variables
3. **ClustVarACM**: Multiple Correspondence Analysis clustering for categorical variables

Each class provides a unified interface with `fit()`, `predict()`, and `plot()` methods.

## Installation

```{r eval=FALSE}
# Install from GitHub
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

remotes::install_github("cyrizon/r-clustering-variables")
```

```{r}
library(clustVarACC)
```

## Example 1: K-Means Variable Clustering

### Basic Usage

Let's cluster numeric variables from the built-in `mtcars` dataset:

```{r kmeans-basic}
# Select numeric variables
data(mtcars)
X <- mtcars[, c("mpg", "disp", "hp", "drat", "wt", "qsec")]

# Create and fit K-means model with 3 clusters
model_km <- ClustVarKMeans$new(K = 3, method = "correlation")
model_km$fit(X)

# View results
print(model_km)
```

### Cluster Composition

```{r kmeans-clusters}
# Access cluster assignments
model_km$clusters
```

### Automatic K Selection

The package includes an elbow method to automatically select the optimal number of clusters:

```{r kmeans-auto-k, eval=FALSE}
# Auto-detect optimal k using elbow method
model_km_auto <- ClustVarKMeans$new(K = 2, method = "correlation")
optimal_k <- model_km_auto$elbow_method(X, K_min = 2, K_max = 8, plot = TRUE)

cat("Optimal K:", optimal_k, "\n")

# Fit with optimal k
model_km_auto$K <- optimal_k
model_km_auto$fit(X)
```

### Visualizations

```{r kmeans-plots, fig.width=7, fig.height=5}
# Correlation heatmap ordered by clusters
model_km$plot(type = "heatmap")

# Variable representativeness (similarity to cluster center)
model_km$plot(type = "representativeness")
```

### Prediction on New Variables

```{r kmeans-predict}
# Create illustrative variables (combinations of existing ones)
new_vars <- data.frame(
    power_weight_ratio = mtcars$hp / mtcars$wt,
    efficiency = mtcars$mpg * mtcars$drat
)

# Predict cluster membership
predictions <- model_km$predict(new_vars)
print(predictions)
```

## Example 2: Hierarchical Clustering (HAC)

HAC is useful when you want to visualize the hierarchical structure of variable relationships.

### Basic Usage

```{r hac-basic}
# Create and fit HAC model
model_hac <- ClustVarHAC$new(K = 3, method = "correlation", linkage_method = "ward.D2")
model_hac$fit(X)

print(model_hac)
```

### Dendrogram Visualization

```{r hac-dendrogram, fig.width=7, fig.height=5}
# Plot dendrogram with cut line
model_hac$plot(type = "dendrogram")
```

### Choosing the Number of Clusters

```{r hac-heights, fig.width=7, fig.height=5}
# Plot fusion heights to identify "elbow"
model_hac$plot(type = "heights")
```

### Additional Visualizations

```{r hac-other-plots, fig.width=7, fig.height=5}
# Heatmap
model_hac$plot(type = "heatmap")

# Representativeness
model_hac$plot(type = "representativeness")
```

## Example 3: ACM Clustering for Categorical Variables

For categorical variables, we use Multiple Correspondence Analysis (MCA) to define cluster axes.

### Basic Usage

```{r acm-basic}
# Load a dataset with categorical variables
data(HairEyeColor)
hair_eye <- as.data.frame(HairEyeColor)

# Expand the contingency table to individual observations
hair_eye_expanded <- hair_eye[rep(seq_len(nrow(hair_eye)), hair_eye$Freq), c("Hair", "Eye", "Sex")]
rownames(hair_eye_expanded) <- NULL

# Ensure all variables are factors
hair_eye_expanded[] <- lapply(hair_eye_expanded, as.factor)

# Create and fit ACM model
model_acm <- ClustVarACM$new(K = 2, max_iter = 30)
model_acm$fit(hair_eye_expanded)

print(model_acm)
```

### Cluster Results

```{r acm-clusters}
# View cluster assignments
model_acm$clusters

# View final quality criterion (sum of chi-squared associations)
model_acm$Q_final
```

### Visualizations

```{r acm-plots, fig.width=7, fig.height=5}
# MCA biplot showing variables and clusters
model_acm$plot(type = "biplot")

# Association heatmap (Cramér's V)
model_acm$plot(type = "heatmap")

# Variable representativeness
model_acm$plot(type = "representativeness")
```

### Convergence Tracking

```{r acm-convergence}
# Check convergence trace
plot(model_acm$Q_trace,
    type = "b",
    xlab = "Iteration", ylab = "Q Criterion",
    main = "ACM Convergence", pch = 19, col = "steelblue"
)
```

## Comparing Distance Methods

For numeric variables, you can choose between correlation and Euclidean distance:

```{r compare-methods, fig.width=7, fig.height=10}
# Correlation-based (captures linear relationships)
model_cor <- ClustVarKMeans$new(K = 3, method = "correlation")
model_cor$fit(X)

# Euclidean-based (captures scale and magnitude)
model_euc <- ClustVarKMeans$new(K = 3, method = "euclidean")
model_euc$fit(X)

# Compare cluster compositions
cat("Correlation method:\n")
print(model_cor$clusters)

cat("\nEuclidean method:\n")
print(model_euc$clusters)
```

## Method Chaining

All models support method chaining for concise code:

```{r method-chaining, eval=FALSE}
# Chain initialization, fitting, and plotting
ClustVarKMeans$new(K = 3, method = "correlation")$
    fit(X)$
    plot(type = "heatmap")
```

## Exporting Results

Extract cluster assignments for further analysis:

```{r export}
# Convert to data.frame
results_df <- data.frame(
    variable = unlist(model_km$clusters),
    cluster = rep(
        seq_along(model_km$clusters),
        sapply(model_km$clusters, length)
    )
)

print(results_df)

# Export to CSV
# write.csv(results_df, "cluster_results.csv", row.names = FALSE)
```

## Shiny Application

The package includes an interactive Shiny application for exploratory analysis without coding:

```{r shiny, eval=FALSE}
# Launch the Shiny app from installed package
library(clustVarACC)
shiny::runApp(system.file("shinyApp", package = "clustVarACC"))
```

**Note**: When running from source (development mode), use the package root directory:

```{r shiny-dev, eval=FALSE}
# Development mode - ensure you're in the project root directory
# Find package root (works from any subdirectory)
pkg_root <- rprojroot::find_package_root_file()
devtools::load_all(pkg_root) # Load package functions
shiny::runApp(file.path(pkg_root, "inst/shinyApp"))

# Alternative if rprojroot is not installed:
# setwd("..") if you're in vignettes/, then:
# devtools::load_all()
# shiny::runApp("inst/shinyApp")
```

The Shiny app provides:

- Data upload and preview
- Interactive algorithm selection (K-Means, HAC, ACM)
- Parameter tuning (K, distance method, linkage)
- Automatic K detection
- Multiple visualization types
- Results export (CSV, PNG)
- Prediction on new variables

## Best Practices

### Choosing an Algorithm

- **K-Means (ClustVarKMeans)**: Fast, good for large datasets, requires specifying K
- **HAC (ClustVarHAC)**: Hierarchical structure, visual K selection via dendrogram
- **ACM (ClustVarACM)**: For categorical variables only

### Choosing Distance Method

- **Correlation**: Focus on linear relationships (ignores scale)
- **Euclidean**: Considers both magnitude and direction (scale-sensitive)

### Preprocessing

- **Numeric data**: Automatically scaled within each algorithm
- **Categorical data**: Must be factors for ACM
- **Missing values**: Not allowed - impute or remove beforehand

### Interpreting Results

1. **Cluster sizes**: Check for balanced vs. unbalanced clusters
2. **Representativeness plot**: Identify most representative variables per cluster
3. **Heatmap**: Verify intra-cluster cohesion (high correlation/association within clusters)
4. **Prediction**: Validate cluster structure by predicting new variable assignments

## Advanced: Custom Linkage Methods (HAC)

```{r hac-linkage, eval=FALSE}
# Try different linkage methods
linkages <- c("ward.D2", "complete", "average", "single")

for (link in linkages) {
    model <- ClustVarHAC$new(K = 3, method = "correlation", linkage_method = link)
    model$fit(X)
    cat("\nLinkage:", link, "\n")
    print(model$clusters)
}
```

## Troubleshooting

### Empty Clusters

If K-means produces empty clusters, try:

- Increasing `nstart` parameter (more random initializations)
- Reducing K
- Using HAC instead

### ACM Convergence Issues

If ACM doesn't converge:

- Increase `max_iter`
- Adjust `tol` (tolerance)
- Check data quality (too many rare categories?)

### Prediction Errors

Ensure new data:

- Has the same number of observations as training data
- Contains only numeric variables (K-Means/HAC) or factors (ACM)
- Has no missing values

## Summary

The **clustVarACC** package provides a modern, R6-based interface for variable clustering with:

- Three complementary algorithms (K-Means, HAC, ACM)
- Unified `fit()`, `predict()`, `plot()` methods
- Automatic K selection (K-Means)
- Rich visualizations (heatmaps, dendrograms, biplots)
- Interactive Shiny application
- Prediction on new variables

For more information, see the package documentation:

```{r help, eval=FALSE}
?ClustVarKMeans
?ClustVarHAC
?ClustVarACM
```

## References

- Vigneau, E., & Qannari, E. M. (2003). Clustering of variables around latent components. *Communications in Statistics - Simulation and Computation*, 32(4), 1131-1150.
- Chavent, M., Kuentz-Simonet, V., Labenne, A., & Saracco, J. (2012). ClustOfVar: An R package for the clustering of variables. *Journal of Statistical Software*, 50(13), 1-16.
